---
title: "STAT 331/531 Final Project: Checkpoint 3 Progress Report"
author: "Elaina Chao, Natalie Murai, Nathan Munoz"
format: html
editor: visual
embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

In this lab, we will analyze Division I college basketball data from the 2013-2026 seasons. The dataset includes team statistics such as wins, losses, offensive efficiency, defensive efficiency, strength of schedule, and tournament-related performance measures.

The goal of this lab is to practice data wrangling, visualization, and functional programming skills using real sports analytics data. This dataset works well for the lab because it contains multiple years of observations, categorical and numeric variables, opportunities for grouping and summarizing, and enough complexity for writing functions and iteration tasks.

## Dataset Source

College Basketball Dataset from Kaggle: <https://www.kaggle.com/datasets/andrewsundberg/college-basketball-dataset/data>

## Data Dictionary: Selected Variables

| Variable  | Description                                        |
|-----------|----------------------------------------------------|
| `TEAM`    | Team name                                          |
| `CONF`    | Conference                                         |
| `YEAR`    | Season year                                        |
| `G`       | Games played                                       |
| `W`       | Wins                                               |
| `ADJOE`   | Adjusted offensive efficiency                      |
| `ADJDE`   | Adjusted defensive efficiency                      |
| `BARTHAG` | Power rating metric                                |
| `EFG_O`   | Effective field goal percentage on offense         |
| `EFG_D`   | Effective field goal percentage allowed on defense |
| `TOR`     | Turnover rate                                      |
| `WAB`     | Wins above bubble                                  |

## Learning Goals

This lab is intended to assess the following skills:

1.  Using `dplyr` verbs for efficient data manipulation.
2.  Creating data visualizations with `ggplot2`.
3.  Writing functions that take user inputs and return data frame outputs.
4.  Using string and data-cleaning tools to prepare data for analysis.
5.  Interpreting statistical graphics and summaries in context.

## Setup

```{r}
library(tidyverse)
library(kableExtra)

basketball <- read_csv("cbb.csv")
```

# Lab Questions

## Question 1: Data Wrangling

Create a new dataset containing only teams from the 2024 season.

Then:

1.  Select the variables `TEAM`, `CONF`, `W`, `ADJOE`, and `ADJDE`.
2.  Arrange the teams from most wins to fewest wins.
3.  Display the first 10 rows.

### Solution

```{r}
basketball_2024 <- basketball |>
  filter(YEAR == 2024) |>
  select(TEAM, CONF, W, ADJOE, ADJDE) |>
  arrange(desc(W))

head(basketball_2024, 10)
```

## Question 2: Visualization

Using the 2024 season data, create a scatterplot showing the relationship between adjusted offensive efficiency (`ADJOE`) and adjusted defensive efficiency (`ADJDE`). Color the points by conference.

Then answer the following questions:

1.  Which teams appear to be strongest overall?
2.  Do certain conferences appear clustered together?

### Solution

```{r}
basketball_2024 |>
  ggplot(aes(x = ADJOE, y = ADJDE, color = CONF)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Offensive vs. Defensive Efficiency in the 2024 Season",
    x = "Adjusted Offensive Efficiency",
    y = "Adjusted Defensive Efficiency",
    color = "Conference"
  ) +
  theme_minimal()
```

Strong teams generally appear with high offensive efficiency and low defensive efficiency. Some conferences may show visible clustering, suggesting similar performance levels among teams in the same conference.

## Question 3: Functions: Week 7+ Material

Write a function called `top_team()` that:

-   takes a conference name as input,
-   filters the dataset to the 2024 season,
-   and returns the team with the highest number of wins in that conference.

Test your function using the `"WCC"` and `"SEC"` conferences.

### Solution

```{r}
top_team <- function(conf_name) {
  basketball |>
    filter(YEAR == 2024, CONF == conf_name) |>
    arrange(desc(W)) |>
    slice(1) |>
    select(TEAM, CONF, W)
}
```

```{r}
top_team("WCC")
top_team("SEC")
```

## Question 4: Data Wrangling

Compare conference performance between the 2013 and 2024 seasons. For each conference, calculate the median adjusted offensive efficiency and median adjusted defensive efficiency in both years.

-   Create one row for each conference.
-   Include columns for 2013 offense, 2024 offense, offense change, 2013 defense, 2024 defense, and defense change.
-   Arrange the table from largest offensive improvement to smallest offensive improvement.
-   Display the first 10 rows.

::: {.callout-tip}
The `median()` function will be helpful for finding the typical `ADJOE` and `ADJDE` values for each conference and year.

For offense, higher is better, so use:

`offense_change = median_offense_2024 - median_offense_2013`

For defense, lower is better, so a negative `defense_change` means the conference improved defensively.
:::

### Solution

```{r}
conference_table <- basketball |>
  filter(YEAR == 2013 | YEAR == 2024) |>
  group_by(CONF, YEAR) |>
  summarize(
    median_offense = median(ADJOE, na.rm = TRUE),
    median_defense = median(ADJDE, na.rm = TRUE)
  ) |>
  pivot_wider(
    names_from = YEAR,
    values_from = c(median_offense, median_defense)
  ) |>
  drop_na() |>
  mutate(
    offense_change = median_offense_2024 - median_offense_2013,
    defense_change = median_defense_2024 - median_defense_2013
  ) |>
  arrange(desc(offense_change)) |>
  head(10)
```

## Question 5: Table Cleaning with KableExtra

Using the table you created in Question 4, make a cleanly formatted table that displays the first 10 rows.

Your table should:

1. Display the top 10 conferences from Question 4.
2. Use `kable_styling()` to improve the appearance of the table.
3. Include a clear title and readable column names.
4. Clean the values to include only two decimal places.

### Solution
```{r}
conference_table |>
  kable(
    caption = "Top 10 Conferences by Offensive Improvement from 2013 to 2024",
    col.names = c(
      "Conference",
      "2013 Offense",
      "2024 Offense",
      "2013 Defense",
      "2024 Defense",
      "Offense Change",
      "Defense Change"
    ),
    digits = 2
  ) |>
  kable_styling(
    full_width = FALSE,
    bootstrap_options = "striped"
  )
```


# Planned Additions for Final Submission

For the final version of the lab, we plan to include simulation-based questions, iteration using `map()`, more advanced visualizations, and challenging multi-step data wrangling problems.

We also plan to add 8--12 total questions, a complete solutions document, and a short written reflection discussing the learning goals and design of the lab.
